Time Series Analysis with Bandt \& Pompe Symbolization and a Test for White Noise in the Entropy-Complexity Plane

### 2021-10-25 Reports/JOURNAL - Confidence Regions/IT-ConfidenceRegionsR1.tex
Update IT-ConfidenceRegionsR1.tex

\begin{abstract}
This article serves two purposes.
Firstly, it surveys the \citeauthor{PermutationEntropyBandtPompe} methodology for the statistical community, stressing topics that are open for research.
Secondly, it contributes towards a better understanding of the statistical properties of that approach for time series analysis.
%
The \citeauthor{PermutationEntropyBandtPompe} methodology 
%has been used successfully in the analysis of time series. 
%It 
consists of computing
Information Theory descriptors of the histogram of ordinal patterns. 
Such descriptors lie in a 2D manifold: the Entropy-Complexity plane. 
This article provides the first proposal
of a test in the Entropy-Complexity plane for the
white noise hypothesis.
Our test is based on true white noise sequences obtained from physical devices.
The proposed methodology provides consistent results, 
managing to assess sequences of true random samples as random (adequate test size), 
rejecting correlated sequences (sound test power), and
capturing the randomness of generators previously analyzed in the literature.
\keywords{Random number generators, Bandt–Pompe approach, Entropy-complexity plane, Time Series, Information Theory}
\end{abstract}

@@

The \num{2,160} citations received by the seminal paper appeared in \num{780} venues indexed by the Web of Science.
Among them, the journals belong to \num{127} categories, spanning from Multidisciplinary Physics (\SI{24}{\percent} of the publications) to Zoology (only one of the of citing articles).
There are \num{22} citing articles from journals that belong to the Statistics \& Probability category.
Five of these articles appeared in \textit{Stochastic Environmental Research and Risk Assessment}, 
two in the \textit{Journal of Time Series Analysis} and in 
\textit{Theory and Applications of Time Series Analysis},
and each of the remaining ten appeared in a different journal.
Most of these articles relate successful applications of the Bandt and Pompe methodology, except \cite{OrdinalPatternProbabilities} that obtained the sample entropy's properties under zero-mean Gaussian processes.
It is also noteworthy that, in this category of publications, \cite{DistributionsofOrderPatternsofIntervalMaps} provided a formal and more general proof of the structure of the boundary of the $H\times C$ manifold than that obtained by \cite{martin2006generalized}.
The lack of attention that the Bandt and Pompe approach has received by the Probability \& Statistics community confirms that it is a fertile research avenue waiting to explore.

@@

With the knowledge of the expected variability of such points, according to the underlying dynamics, we can make hypothesis tests for a wide variety of models.
Results in this direction can be found in the literature.
\cite{RandomNumberGeneratorsCausality} showed that the Entropy-Complexity plane ($H\times C$) is a good indicator of Diehard tests' results on pseudorandom number generators.
\cite{De_Micco_2008} assessed ways of improving pseudorandom sequences by their representation in this plane.

Motivated by previous works, in this paper, we advance the state-of-the-art providing the first test for white noise points in the $H\times C$ plane.
In this proposal, the input is a sequence of true random observations generated by a physical-based procedure. 
We obtain the confidence regions by performing an orthogonal projection of the data onto the space of principal components, thus eliminating the restrictions imposed by the bounded space of the Entropy-Complexity plane.
Our contributions can be summarized as follows:
\begin{itemize}
	\item We provide the first contribution in constructing a test in the Entropy-Complexity Plane: we provide confidence regions and $p$-values.
	\item We evaluate this test's size by analyzing random sequences generated by physical procedures and pseudorandom generators (PRNGs).
	\item We verify the test's power contrasting correlated noise time series, and white noise series patched with a deterministic signal.
\end{itemize}

@@

In our work, we consider the ordinal patterns formed from true white noise sequences (TWNS) using the Bandt and Pompe symbolization.
Such sequences are then mapped in the two-dimensional plane of Information Theory descriptors, formed by the Permutation Entropy and the Statistical Complexity.
We then obtain a test statistic, using confidence regions, that can discriminate among times series.

@@

The Jensen-Shannon distance between $\bm h$ and the probability function $\bm u=(u(1), u(2), \dots, u(D!))$ stems as a more adequate measure of such a distance:
\begin{equation}
	Q'(\bm{h}, \bm{u}) = \sum_{\ell=1}^{D!} \Big(h(\ell) \log\frac{h(\ell)}{u(\ell)} +
	u(\ell) \log\frac{u(\ell)}{h(\ell)}
	\Big).
	\label{eq:JensenShannon}
\end{equation}
This has been the preferred distance function since its proposal by \citet{Lamberti2004}.
Most works consider the uniform distribution $\bm u = (1/D!, 1/D!, \dots, 1/D!)$ as the reference model.

@@

Fig.~\ref{fig:RightMostCorner} shows the rightmost lower corner of the $H\times C$ plane, emphasizing the location of the white ($k=0$), $k=1/2$, and pink ($k=1$) noises.
In particular, notice that the point associated to white noise is not at $(1,0)$, as consequence of being obtained from a finite rather than infinite length sequence.

@@

Due to the infinitude of white noise sequences, although these sequences have the characteristic of presenting high values of entropy and low statistical complexity, their points will not necessarily be located in $(1, 0)$, but in a surrounding region.
Our study's focus is to assess pure randomness by analyzing the empirical distribution of the points produced by true random sequences of finite size and obtaining regions of confidence in the $H \times C$ plane.

@@

None of these works provide $p$-values or hypothesis tests for their analysis.
The focus of the articles is finding the set of descriptors that best discriminates chaos from noise.
The authors make assessments about the randomness of sequence on ad hoc visual inspection of the point's location in the $H\times C$ plane.
Our work fills such a gap for finite sequences of white noise and proposes a methodology that can be extended to any other situation.

@@

In this section, we formalize the task of building confidence regions in the Entropy-Complexity manifold.
Then, we present our proposal to change space through the algorithm of the principal components analysis.
Our goal is to find a latent space representative of the data without the restrictions of the $H\times C$ plane's boundaries.
Through this new representation of the data, we calculate empirical regions with different levels of confidence.
Finally, after calculating these regions, we build a test statistic that determines the probability that a given sequence belongs to the distribution of the points provided.

@@

Random numbers are used in many fields, from gambling to cryptography, to guarantee a secure, realistic, or unpredictable behavior. 
Pseudorandom results can be achieved by software in a deterministic way, but some applications need actual random numbers (despite the somewhat elusive nature of actual randomness).
Randomness can be observed in unpredictable real-world phenomena like cathodic radiation or atmospheric noise.
%In this study we used two sources of real random numbers. 
%The first is based on vacuum states to generate random quantum numbers described by \citeauthor{RNGVacuumStates}~(\citeyear{RNGVacuumStates}), the second one is based on atmospheric noise captured by a cheap radio receiver presented at \url{www.random.org}.

In this study, we used two sources of true random numbers, both from the observation and measurement of physical phenomena.
The first uses vacuum states. 
The setup consists of an ordinary laser source to generate a local oscillator (LO), a half-wave plate, a polarizing beamsplitter (BPS), and two balanced detectors working together adding or subtracting the photocurrents results in a quadrature measurement of the LO or vacuum state. 
The distribution of the vacuum state is binned into $2^n$ equal parts (bins of the same size), assigning a fixed bit combination of length $n$ to each sample point in a given bin \citep{RNGVacuumStates}. 
The second one employs atmospheric noise captured by a cheap radio receiver with no filter for unwanted static sounds caused by atmospheric noise.
This generator was implemented over a distributed setup with radios located at different geographical locations sending random bits to a cloud server that processes data and hosts the values.
The history of this service and other information can be found at \citet{RandomOrg}.

@@

We conducted an ablation study to identify the influence of the parameters $T$, $D$, and $\tau$ in the construction of empirical confidence regions.
We verified that the results involving the time delay parameter variation did not show significant differences in repeated experiments; therefore, in the sequel, we did no consider $\tau$ as a determining factor.
On the other hand, we found two relevant variables: 
the length of the sequence 
and the embedding dimension.
We, thus, employed the following factors:
\begin{itemize}
	\item Sequence length $T\in\mathcal T=\{ \num[scientific-notation=true]{e3}, \num[scientific-notation=true]{5 e4}\}$,
	\item Embedding dimension $D\in\mathcal D=\{3, 4, 5, 6\}$.
	%,and
	%\item Time delay $\tau=\{1, 10, 30, 50\}$.
\end{itemize}
and kept $\tau=1$, which is the most frequently used option.
The values of $D$ are within the range recommended in the literature~\citep{PermutationEntropyBandtPompe}.

@@

The procedure operates on the principal components space and consists of measuring the closeness between the ``emblematic point'' and the observed point.
We are given a time series $\bm x$ of size $T$, and we want its $p$-value when contrasted with TWNRS of the same size at embedding dimension $D$.
We use $N$ TWNRS of size $T$, compute their points in the $H\times C$ plane, and project them to the corresponding principal components space.
We then do the same with $\bm x$, and obtain a new point $(u_x,v_x)$.
The closer $\bm x$ is to the emblematic time series, the larger its $p$-value.
Assume that the emblematic time series is represented by $(u,v)$ in the principal components space.
We measure this closeness by building a box around $(u_x,v_x)$ that contains $(u,v)$; assume that $u_x>u$, then:
\begin{enumerate}
	\item the right side of the box is the smallest $u_j$ which is larger that $u_x$; assume it corresponds to the quantile $\eta_u$ of $\utilde u = (u_1,u_2,\dots, u_N)$. By definition, $\eta_u\geq 1/2$.
	%
	\item the left side of the box is the $1-\eta_u$ quantile of $\utilde u$.
	%
	\item the top side of the box is the smallest $v_j$ which is larger that $v_x$; assume it corresponds to the quantile $\eta_v$ of $\utilde v = (v_1,v_2,\dots, v_N)$. By definition, $\eta_v\geq 1/2$.
	%
	\item the bottom side of the box is the $1-\eta_v$ quantile of $\utilde v$.
\end{enumerate}
Fig.~\ref{fig:methodologyPvalue} illustrates theses steps.

@@

Fig.~\ref{fig:HC-PCA} shows the results for $T = 50000$ and $D = 3,6$ in the new principal components space, along with the quantiles of order $\SI{90}{\percent}$, $\SI{95}{\percent}$, $\SI{99}{\percent}$, and $\SI{99.9}{\percent}$.
We also show the projection of the $H \times C$ plane boundaries in this space, as well the median of each data set, the latter being represented as red dots.

@@

Fig.~\ref{fig:HC-PCA} also shows that the data are not evenly distributed among the axes of the first principal component.
They tend to concentrate close to the point that corresponds to $(1,0)$ in the $H\times C$ plane.
As we use order statistics to define the confidence regions, this issue is also of little relevance for our results.
Moreover, Fig.~\ref{fig:PCA-Hist} shows that such asymmetry diminishes when the embedding dimension $D$ increases.

@@

We assessed the power of the test by contrasting time series with different correlation structure (under the $f^{-k}$ model) in the $H \times C$ plane.
Several studies in the literature have used this approach for identifying and characterizing randomness.

@@

Fig.~\ref{fig:CorrNoisea} shows the overall effect of transforming the emblematic time series into $f^{-k}$ correlated noise, with $k=1/2,1,3/2,2,5/2,3$.
At this scale, the emblematic time series $k=0$ and the one with $k=1/2$ appear overlapped.
As the correlation increases with $k$, the randomness decreases, causing a drop in the entropy; the series become progressively more predictable.

@@

Consider the times series of length $T$ comprised of $T-[c100]$ white noise observations $x_1,x_2,\dots, x_{T-[c100]}$ followed by the signal $y_{T-[c100]+1}, y_{T-[c100]+2}, \dots, y_T$, in which $0\leq c \ll 1$ is the percentage of contamination.
We will verify the behavior of our test for the case $T=1000$, and $y_t$ an increasing function on $t$.
Fig.~\ref{Fig:PointsPatchedIncreasingFunction} shows the effect of such a contamination on the white noise time series whose point in the $H\times C$ plane appears in black.
The Figure also shows the confidence region at \SI{99}{\percent}.

@@

The original point is displaced, at all levels of contamination, to a region with smaller entropy and larger complexity.
When the percentage of contamination is below approximately \SI{2}{\percent}, the test will not reject the null hypothesis at the \SI{99}{\percent}; in fact, the $p$-value is of around \SI{1.5}{\percent} for $c=0,0.01,0.02$.
The $p$-value falls to orders of $10^{-6}$ when the contamination is of \SI{3}{\percent} and above.

@@

\subsection{Revisiting the White Noise Hypothesis in the Literature}

In this section, we compare the performance of our test with that of 
previous analyses that employ the Entropy-Complexity plane.
To this aim, we produced $100$ sequences of length $T = \num[scientific-notation=true]{5 e4}$ for each generator and computed the $p$-value for each $D = \{3, 4, 5, 6\}$.
Previous results are shown in Table~\ref{Tab:Literature}, and ours are in Table~\ref{Tab:LiteratureComparations}.
We grouped our results in those that rejected (R) the null hypothesis and those that did not reject it (NR).

\begin{table}
	\caption{Results of the sequences generated by the main PRNGs in the literature. 
		The sequences have length $T=\num[scientific-notation = true]{5 e4}$.}
	\label{Tab:LiteratureComparations}
	\centering
	\begin{tabular}{cccc}
		\toprule
		Algorithm & 
		\multicolumn{1}{c}{$D$} & 
		$p$-value &
		HC-PCA\\
		\cmidrule(lr){1-1}
		\cmidrule(lr){2-2}
		\cmidrule(lr){3-3}
		\cmidrule(lr){4-4}
		MOT & 3 & 0.305 & NR\\
		& 4 & 0.572 & NR\\ 
		& 5 & 0.455 & NR\\ 
		& 6 & 0.508 & NR\\ 
		\cmidrule(lr){1-4}
		MWC & 3 & 0.501 & NR\\
		& 4 & 0.477 & NR\\ 
		& 5 & 0.496 & NR\\ 
		& 6 & 0.496 & NR\\ 
		\cmidrule(lr){1-4}
		COM & 3 & 0.123 & NR\\
		& 4 & 0.002 & R\\ 
		& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		\cmidrule(lr){1-4}
		LEH & 3 & 0.531 & NR\\
		& 4 & 0.515 & NR\\ 
		\bottomrule
	\end{tabular}
	\begin{tabular}{|cccc}
		\toprule
		Algorithm & 
		\multicolumn{1}{c}{$D$} & 
		$p$-value &
		HC-PCA\\
		\cmidrule(lr){1-1}
		\cmidrule(lr){2-2}
		\cmidrule(lr){3-3}
		\cmidrule(lr){4-4}
		LEH & 5 & 0.495 & NR\\ 
		& 6 & 0.501 & NR\\ 
		\cmidrule(lr){1-4}
		fGn & 3 & 0.521 & NR\\
		& 4 & 0.519 & NR\\ 
		& 5 & 0.498 & NR\\ 
		& 6 & 0.470 & NR\\
		\cmidrule(lr){1-4}
		%fBm & 3 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 4 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%\cmidrule(lr){1-4}
		$f^{-k}$ & 3 & 0.482 & NR\\
		& 4 & 0.520 & NR\\ 
		& 5 & 0.513 & NR\\ 
		& 6 & 0.508 & NR\\
		\cmidrule(lr){1-4}
		LCG & 3 & 0.009 & R\\ 
		& 4 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		\bottomrule
	\end{tabular}
\end{table}

@@

We presented and evaluated the first test for white noise in the  Entropy-Complexity plane.
Our proposal is based on two stages:
(1)~building nonparametric empirical confidence regions in the principal components space and mapping these boxes back to the $H\times C$ plane.
(2)~computing an approximate $p$-value for a given sample by comparing it with points produced by true white noise random sequences (TWNRS).
We obtained the TWNRS with data from physical devices.

Our test has a good size, mostly with long TWNRS.
We also determined the power of our test for the alternative hypothesis of correlated $f^{-k}$ noise and found that it rejects the null hypothesis ($k=0$) for $k>3/10$.

Although our work focuses on the study of short sequences, we were able to capture the random behavior of well-known pseudorandom number generators already analyzed in the literature. 
With this, we verified the adequacy of our technique as it is capable of detecting correlation structures.

>>>>

\begin{abstract}
	This article serves two purposes.
	Firstly, it surveys the \citeauthor{PermutationEntropyBandtPompe} methodology for the statistical community, stressing topics that are open for research.
	Secondly, it contributes towards a better understanding of the statistical properties of that approach for time series analysis.
	%
	The \citeauthor{PermutationEntropyBandtPompe} methodology 
	%has been used successfully in the analysis of time series. 
	%It 
	consists of computing
	Information Theory descriptors of the histogram of ordinal patterns. 
	Such descriptors lie in a 2D manifold: the Entropy-Complexity plane. 
	This article provides the first proposal
	of a test in the Entropy-Complexity plane for the
	white noise hypothesis.
	Our test is based on true white noise sequences obtained from physical devices.
	The proposed methodology provides consistent results: 
	it assesses sequences of true random samples as random (adequate test size), 
	rejects correlated and contaminated sequences (sound test power), and
	captures the randomness of generators previously analyzed in the literature.
	\keywords{Random number generators, Bandt–Pompe approach, Entropy-complexity plane, Time Series, Information Theory}
\end{abstract}

@@

The seminal paper has received \num{2,160} citations from articles published in \num{780} venues indexed by the Web of Science.
Among them, the journals belong to \num{127} categories, spanning from Multidisciplinary Physics (\SI{24}{\percent} of the publications) to Zoology (only one of the citing articles).
There are \num{22} citing articles from journals that belong to the Statistics \& Probability category.
Five of these articles appeared in \textit{Stochastic Environmental Research and Risk Assessment}, 
two in the \textit{Journal of Time Series Analysis} and in 
\textit{Theory and Applications of Time Series Analysis}.
Furthermore, each of the remaining ten appeared in a different journal.
Most of these articles relate successful applications of the Bandt and Pompe methodology, except \cite{OrdinalPatternProbabilities} that obtained the sample entropy's properties under zero-mean Gaussian processes.
It is also noteworthy that, in this category of publications, \cite{DistributionsofOrderPatternsofIntervalMaps} provided a formal and more general proof of the structure of the boundary of the $H\times C$ manifold than that obtained by \cite{martin2006generalized}.
The lack of attention that the Bandt and Pompe approach has received by the Probability \& Statistics community confirms that it is a fertile research avenue waiting to explore.

@@

With the knowledge of the expected variability of such points, we can make hypotheses tests for a wide variety of models according to the underlying dynamics.
Results in this direction can be found in the literature.
\cite{RandomNumberGeneratorsCausality} showed that the Entropy-Complexity plane ($H\times C$) is a good indicator of Diehard tests' results on pseudorandom number generators.
\cite{De_Micco_2008} assessed ways of improving pseudorandom sequences by their representation in this plane.

Motivated by previous works, in this paper, we advance the state-of-the-art providing the first test for white noise points in the $H\times C$ plane.
We used sequences of true random observations generated by a physical-based procedure to produce clouds of points. 
We then project these points onto their principal components, find the regions in this representation, and transform them back to the Entropy-Complexity plane.
In this way, we circumvent the restrictions imposed by this 2D bounded manifold.
We also provide an algorithm that, using geometrical arguments, computes the empirical $p$-value of a sequence under the white noise null hypothesis.
Our contributions can be summarized as follows:
\begin{itemize}
	\item We provide the first contribution in constructing a test in the Entropy-Complexity Plane: we provide confidence regions and $p$-values.
	\item We evaluate this test's size by analyzing random sequences generated by physical procedures and pseudorandom generators (PRNGs).
	\item We verify the test's power contrasting (i)~correlated noise time series and (ii)~white noise series patched with a deterministic signal.
\end{itemize}

@@

Our work considers the ordinal patterns formed from true white noise sequences (TWNS) using the Bandt and Pompe symbolization.
Such sequences are then mapped in the two-dimensional plane of Information Theory descriptors, formed by the Permutation Entropy and the Statistical Complexity.
We then obtain a test statistic, using confidence regions, that can discriminate among times series.

@@

The Jensen-Shannon distance between $\bm h$ and the probability function $\bm u=(u(1), u(2), \dots, u(D!))$ stems as an adequate measure of such a distance:
\begin{equation}
	Q'(\bm{h}, \bm{u}) = \sum_{\ell=1}^{D!} \Big(h(\ell) \log\frac{h(\ell)}{u(\ell)} +
	u(\ell) \log\frac{u(\ell)}{h(\ell)}
	\Big).
	\label{eq:JensenShannon}
\end{equation}
This has been the preferred distance function since its proposal by \citet{Lamberti2004}.
Most works consider the uniform distribution $\bm u = (1/D!, 1/D!, \dots, 1/D!)$ as the reference model.

@@

Fig.~\ref{fig:RightMostCorner} shows the rightmost lower corner of the $H\times C$ plane, emphasizing the location of the white ($k=0$), $k=1/2$, and pink ($k=1$) noises.
In particular, notice that the point associated with white noise is not at $(1,0)$, resulting from a finite rather than infinite length sequence.

@@

Due to the infinitude of white noise sequences, although these sequences have the characteristic of presenting high entropy values and low statistical complexity, their points will not necessarily be located in $(1, 0)$, but in a surrounding region.
Our study focuses on assessing pure randomness by analyzing the empirical distribution of the points produced by true random sequences of finite size and obtaining regions of confidence in the $H \times C$ plane.

@@

None of these works provide $p$-values or hypothesis tests for their analysis.
The focus of the articles is finding the set of descriptors that best discriminates chaos from noise.
The authors assess the randomness of sequences by ad hoc visual inspection of the point's location in the $H\times C$ plane.
Our work fills such a gap for finite sequences of white noise and proposes a methodology that can be extended to any other situation.

@@

In this section, we formalize the task of building confidence regions in the Entropy-Complexity manifold.
Then, we present our proposal to change space through the algorithm of the principal components analysis.
Our goal is to find a space to represent the data without the $H\times C$ plane's boundaries restrictions.
Through this new representation of the data, we calculate empirical regions with different levels of confidence.
Finally, after calculating these regions, we build a test statistic that determines the probability that a given sequence belongs to the distribution of the points provided.

@@

Random numbers are used in many fields, from gambling to cryptography, to guarantee a secure, realistic, or unpredictable behavior. 
Software can produce pseudorandom sequeces in a deterministic way, but some applications need actual random numbers (despite the somewhat elusive nature of actual randomness).
Randomness can be observed in unpredictable real-world phenomena like cathodic radiation or atmospheric noise.
%In this study we used two sources of real random numbers. 
%The first is based on vacuum states to generate random quantum numbers described by \citeauthor{RNGVacuumStates}~(\citeyear{RNGVacuumStates}), the second one is based on atmospheric noise captured by a cheap radio receiver presented at \url{www.random.org}.

In this study, we used two sources of true random numbers, both from the observation and measurement of physical phenomena.
The first uses vacuum states. 
The setup consists of an ordinary laser source to generate a local oscillator (LO), a half-wave plate, a polarizing beamsplitter (BPS), and two balanced detectors, adding or subtracting the photocurrents results in a quadrature measurement of the LO or vacuum state. 
The distribution of the vacuum state is binned into $2^n$ equal parts (bins of the same size), assigning a fixed bit combination of length $n$ to each sample point in a given bin \citep{RNGVacuumStates}. 
The second one employs atmospheric noise captured by a cheap radio receiver with no filter for unwanted static sounds caused by atmospheric noise.
This generator was implemented over a distributed setup with radios at different geographical locations sending random bits to a cloud server that processes data and hosts the values.
The history of this service and other information can be found at \citet{RandomOrg}.

@@

We conducted an ablation study to identify the influence of the parameters $T$, $D$, and $\tau$ in constructing empirical confidence regions.
We verified that the results involving the time delay parameter variation did not show significant differences in repeated experiments; therefore, we did not consider $\tau$ as a determining factor in the sequel.
On the other hand, we found two relevant variables: 
the length of the sequence 
and the embedding dimension.
We, thus, employed the following factors:
\begin{itemize}
	\item Sequence length $T\in\mathcal T=\{ \num[scientific-notation=true]{e3}, \num[scientific-notation=true]{5 e4}\}$,
	\item Embedding dimension $D\in\mathcal D=\{3, 4, 5, 6\}$.
	%,and
	%\item Time delay $\tau=\{1, 10, 30, 50\}$.
\end{itemize}
and kept $\tau=1$, which is the most frequently used option.
The values of $D$ are within the range recommended in the literature~\citep{PermutationEntropyBandtPompe}.

@@

The procedure operates on the principal components space and measures the closeness between the ``emblematic point'' and the observed point.
We are given a time series $\bm x$ of size $T$, and we want its $p$-value when contrasted with TWNRS of the same size at embedding dimension $D$.
We use $N$ TWNRS of size $T$, compute their points in the $H\times C$ plane, and project them to the corresponding principal components space.
We then do the same with $\bm x$, and obtain a new point $(u_x,v_x)$.
The closer $\bm x$ is to the emblematic time series, the larger its $p$-value.
Assume that the emblematic time series is represented by $(u,v)$ in the principal components space.
We measure this closeness by building a box around $(u_x,v_x)$ that contains $(u,v)$; assume that $u_x>u$, then:
\begin{enumerate}
	\item the right side of the box is the smallest $u_j$ which is larger that $u_x$; assume it corresponds to the quantile $\eta_u$ of $\utilde u = (u_1,u_2,\dots, u_N)$. By definition, $\eta_u\geq 1/2$.
	%
	\item the left side of the box is the $1-\eta_u$ quantile of $\utilde u$.
	%
	\item the top side of the box is the smallest $v_j$ which is larger that $v_x$; assume it corresponds to the quantile $\eta_v$ of $\utilde v = (v_1,v_2,\dots, v_N)$. By definition, $\eta_v\geq 1/2$.
	%
	\item the bottom side of the box is the $1-\eta_v$ quantile of $\utilde v$.
\end{enumerate}
Fig.~\ref{fig:methodologyPvalue} illustrates theses steps.

@@

Fig.~\ref{fig:HC-PCA} shows the results for $T = 50000$ and $D = 3,6$ in the new principal components space, along with the quantiles of order $\SI{90}{\percent}$, $\SI{95}{\percent}$, $\SI{99}{\percent}$, and $\SI{99.9}{\percent}$.
We also show the projection of the $H \times C$ plane boundaries in this space and the median of each data set, the latter represented as red dots.

@@

Fig.~\ref{fig:HC-PCA} also shows that the data are not evenly distributed among the axes of the first principal component.
They tend to concentrate close to $(1,0)$ in the $H\times C$ plane.
As we use order statistics to define the confidence regions, this issue is also of little relevance for our results.
Moreover, Fig.~\ref{fig:PCA-Hist} shows that such asymmetry diminishes when the embedding dimension $D$ increases.

@@

We assessed the power of the test by contrasting time series with different correlation structures (under the $f^{-k}$ model) in the $H \times C$ plane.
Several studies in the literature have used this approach for identifying and characterizing randomness.

@@

Fig.~\ref{fig:CorrNoisea} shows the overall effect of transforming the emblematic time series into $f^{-k}$ correlated noise, with $k=1/2,1,3/2,2,5/2,3$.
At this scale, the emblematic time series $k=0$ and the one with $k=1/2$ appear overlapped.
As the correlation increases with $k$, the randomness decreases, causing a drop in the entropy; the series becomes progressively more predictable.

@@

Consider the times series of length $T$ comprised of $T-[c100]$ white noise observations $x_1,x_2,\dots, x_{T-[c100]}$ followed by the signal $y_{T-[c100]+1}, y_{T-[c100]+2}, \dots, y_T$, in which $0\leq c \ll 1$ is the percentage of contamination.
We will verify the behavior of our test for the case $T=1000$, and $y_t$ an increasing function on $t$.
Fig.~\ref{Fig:PointsPatchedIncreasingFunction} shows the effect of such contamination on the white noise time series whose point in the $H\times C$ plane appears in black.
The Figure also shows the confidence region at \SI{99}{\percent}.

@@

At all levels of contamination, the original point is displaced to a region with smaller entropy and larger complexity.
When the percentage of contamination is below approximately \SI{2}{\percent}, the test will not reject the null hypothesis at the \SI{99}{\percent}; in fact, the $p$-value is of around \SI{1.5}{\percent} for $c=0,0.01,0.02$.
The $p$-value falls to orders of $10^{-6}$ when the contamination is of \SI{3}{\percent} and above.

@@

\subsection{Revisiting the White Noise Hypothesis in the literature}

In this section, we compare the performance of our test with that of 
previous analyses that employ the Entropy-Complexity plane.
To this aim, we produced $100$ sequences of length $T = \num[scientific-notation=true]{5 e4}$ for each generator and computed the $p$-value for each $D = \{3, 4, 5, 6\}$.
Previous results are shown in Table~\ref{Tab:Literature}, and ours are in Table~\ref{Tab:LiteratureComparations}.
We grouped our results in those that rejected (R) the null hypothesis and those that did not reject it (NR), given the $p$-value obtained by our test. 
This experiment investigates the robustness of the confidence regions generated by TRWNS in samples from different randomness generators.

\begin{table}
	\caption{Results of the sequences generated by the main PRNGs in the literature. 
		The sequences have length $T=\num[scientific-notation = true]{5 e4}$. 
		The ``Decision'' of rejecting (R) or not (NR) is at the $\SI{5}{\percent}$ significance level.}
	\label{Tab:LiteratureComparations}
	\centering
	\begin{tabular}{cccc}
		\toprule
		Algorithm & 
		\multicolumn{1}{c}{$D$} & 
		$p$-value &
		Decision\\
		\cmidrule(lr){1-1}
		\cmidrule(lr){2-2}
		\cmidrule(lr){3-3}
		\cmidrule(lr){4-4}
		MOT & 3 & 0.305 & NR\\
		& 4 & 0.572 & NR\\ 
		& 5 & 0.455 & NR\\ 
		& 6 & 0.508 & NR\\ 
		\cmidrule(lr){1-4}
		MWC & 3 & 0.501 & NR\\
		& 4 & 0.477 & NR\\ 
		& 5 & 0.496 & NR\\ 
		& 6 & 0.496 & NR\\ 
		\cmidrule(lr){1-4}
		COM & 3 & 0.123 & NR\\
		& 4 & 0.002 & R\\ 
		& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		\cmidrule(lr){1-4}
		LEH & 3 & 0.531 & NR\\
		& 4 & 0.515 & NR\\ 
		\bottomrule
	\end{tabular}
	\begin{tabular}{|cccc}
		\toprule
		Algorithm & 
		\multicolumn{1}{c}{$D$} & 
		$p$-value &
		Decision\\
		\cmidrule(lr){1-1}
		\cmidrule(lr){2-2}
		\cmidrule(lr){3-3}
		\cmidrule(lr){4-4}
		LEH & 5 & 0.495 & NR\\ 
		& 6 & 0.501 & NR\\ 
		\cmidrule(lr){1-4}
		fGn & 3 & 0.521 & NR\\
		& 4 & 0.519 & NR\\ 
		& 5 & 0.498 & NR\\ 
		& 6 & 0.470 & NR\\
		\cmidrule(lr){1-4}
		%fBm & 3 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 4 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		%\cmidrule(lr){1-4}
		$f^{-k}$ & 3 & 0.482 & NR\\
		& 4 & 0.520 & NR\\ 
		& 5 & 0.513 & NR\\ 
		& 6 & 0.508 & NR\\
		\cmidrule(lr){1-4}
		LCG & 3 & 0.009 & R\\ 
		& 4 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 5 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		& 6 & \num[scientific-notation=true]{1.11 e-16} & R\\ 
		\bottomrule
	\end{tabular}
\end{table}

@@

We presented and evaluated the first test for white noise in the  Entropy-Complexity plane.
Our proposal is based on two stages:
(1)~building non-parametric empirical confidence regions in the principal components space and mapping these boxes back to the $H\times C$ plane.
(2)~computing an approximate $p$-value for a given sample by comparing it with points produced by true white noise random sequences (TWNRS).
We obtained the TWNRS with data from physical devices.

Our test has good size, mostly with long TWNRS.
We also determined the power of our test for the alternative hypothesis of correlated $f^{-k}$ noise and found that it rejects the null hypothesis ($k=0$) for $k>3/10$.

Although our work focuses on the study of short sequences, we captured the random behavior of well-known pseudorandom number generators already analyzed in the literature. 
With this, we verified the adequacy of our technique as it is capable of detecting correlation structures.
